Standard Candle Cosmology With Incomplete Spectroscopic Classification

### 2015-09-23 doc/article.tex
start article

\documentclass[preprint]{aastex}
\usepackage{amsmath,amssymb}
\usepackage{mathrsfs}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{bm}
\bibliographystyle{apj}

@@

\title{Model for DES Type Ia Supernova Cosmology Analysis}
\author{Alex Kim}

@@

\section{Simple Example}
\subsection{Model}
\label{example:sec}
A simplified version of the above model,
\begin{equation}
p(\mathit{ADU}, {{T}}_S,{{z}}_S, \theta_G|  \Omega_M, w, \theta_r,\alpha_{Ia},\sigma_{Ia}, \alpha_{\mathit{non-Ia}},\sigma_{\mathit{non-Ia}}),
\end{equation}
is examined in this section.
The measurements: 
\begin{itemize}
\item $\mathit{ADU}$ are measured counts at peak brightness.
\item ${{T}}_S$, ${{z}}_S$ are spectroscopic type and redshift for a subset of transients.
\item $\theta_G$ are the redshift of the galaxy identified as the transient host.
\end{itemize}
The model parameters:
\begin{itemize}
\item $\Omega_M$, $w$ are the parameters of a flat dark-energy cosmology.
\item $\theta_r$ describe the fraction of transients that are SN~Ia, independent
of host-galaxy properties.
\item $\alpha_{Ia}$, $\sigma_{Ia}$, $\alpha_{\mathit{non-Ia}}$, $\sigma_{\mathit{non-Ia}}$ are
the luminosity (times $4\pi$) and intrinsic dispersion for SNe~Ia and non-Ia at peak.
Light-curve shape and color are not considered in this simple model.
\end{itemize}

In addition to the above, there are latent model parameters for each transient used internally
by the model:
\begin{itemize}
\item $z$ are the cosmological redshifts.
\item $T$ are the type, SN~Ia or non-Ia.
\end{itemize}

@@

\subsection{Analysis of Simulated Data}
The simulation of data from the model presented in \S\ref{example:sec}
and their analysis is implemented in Python and are available
on GitHub\footnote{\url{git@github.com:AlexGKim/abc.git}}.

@@

\begin{table}
\centering
\begin{tabular}{|c|c|ccc|}
\hline
Fraction Spec&\%-ile & $w$ &$w$ range & $\Delta w$\\ \hline
$1.00$ &$0.68$ &$-1.08$ & $[-1.196, -0.914]$ & $0.281$ \\
$1.00$ &$0.90$ &$-1.08$ & $[-1.292, -0.826]$ & $0.466$ \\
$1.00$ &$0.95$ &$-1.08$ & $[-1.340, -0.784]$ & $0.556$ \\
$0.60$ &$0.68$ &$-1.08$ & $[-1.227, -0.933]$ & $0.293$ \\
$0.60$ &$0.90$ &$-1.08$ & $[-1.331, -0.836]$ & $0.495$ \\
$0.60$ &$0.95$ &$-1.08$ & $[-1.385, -0.787]$ & $0.598$ \\
$0.20$ &$0.68$ &$-1.15$ & $[-1.249, -0.943]$ & $0.305$ \\
$0.20$ &$0.90$ &$-1.15$ & $[-1.354, -0.851]$ & $0.502$ \\
$0.20$ &$0.95$ &$-1.15$ & $[-1.408, -0.810]$ & $0.597$ \\
$0.00$ &$0.68$ &$-1.09$ & $[-1.218, -0.909]$ & $0.309$ \\
$0.00$ &$0.90$ &$-1.09$ & $[-1.320, -0.812]$ & $0.508$ \\
$0.00$ &$0.95$ &$-1.09$ & $[-1.372, -0.770]$ & $0.602$ \\
\hline
\end{tabular}
\caption{Mode and credible intervals of $w$ for Model 1 for varying
fractions of spectroscopic coverage. \label{seed2:tab}}
\end{table}

@@

When there is 100\% spectroscopic coverage the   $\Omega_M$--$w$  posterior accommodates the input 
parameters, as shown  in
Figure~\ref{contour1:fig}.
A salient feature that contrasts with Model 1 is 
seen in the modes for $w$ listed in Table~\ref{seed2_pop2:tab}: biases
larger than the statistical uncertainties appear with decreasing
spectroscopic coverage.  Note that the Hubble diagrams
in these analyses are almost identical: there are redshift misassignments
for 2\% of the transients that lack spectroscopy.

@@

\begin{table}
\centering
\begin{tabular}{|c|c|ccc|}
\hline
Fraction Spec&\%-ile &$w$ & $w$ range & $\Delta w$\\ \hline
$1.00$ &$0.68$ &$-0.94$ & $[-1.078, -0.804]$ & $0.274$ \\
$1.00$ &$0.90$ &$-0.94$ & $[-1.168, -0.720]$ & $0.447$ \\
$1.00$ &$0.95$ &$-0.94$ & $[-1.214, -0.680]$ & $0.534$ \\
$0.60$ &$0.68$ &$-1.09$ & $[-1.182, -0.883]$ & $0.298$ \\
$0.60$ &$0.90$ &$-1.09$ & $[-1.285, -0.791]$ & $0.494$ \\
$0.60$ &$0.95$ &$-1.09$ & $[-1.332, -0.749]$ & $0.583$ \\
$0.20$ &$0.68$ &$-1.09$ & $[-1.247, -0.936]$ & $0.311$ \\
$0.20$ &$0.90$ &$-1.09$ & $[-1.359, -0.840]$ & $0.518$ \\
$0.20$ &$0.95$ &$-1.09$ & $[-1.413, -0.798]$ & $0.614$ \\
$0.00$ &$0.68$ &$-1.11$ & $[-1.266, -0.948]$ & $0.317$ \\
$0.00$ &$0.90$ &$-1.11$ & $[-1.375, -0.850]$ & $0.525$ \\
$0.00$ &$0.95$ &$-1.11$ & $[-1.428, -0.808]$ & $0.621$ \\
\hline
\end{tabular}
\caption{Mode and credible intervals of $w$ for Model 2  for varying
fractions of spectroscopic coverage.\label{seed2_pop2:tab}}
\end{table}

Some initial conclusions.
\begin{itemize}
\item Knowing the underlying model that produces data, you should
get results statistically consistent with the truth.  A tautology?
\item Typing is important when the non-Ia model is not accurate.
\end{itemize}

@@

\bibliography{/Users/akim/Documents/alex}

>>>>

\documentclass[preprint]{elsarticle}
\usepackage{aas_macros}
\usepackage{amsmath,amssymb}
\usepackage{mathrsfs}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{hyperref}

@@

\begin{frontmatter}

\title{Standard Candle Cosmology With Incomplete Spectroscopic Classification}
\author{A.~G. Kim\corref{cor1}}
\ead{agkim@lbl.gov}
\address{Physics Division, Lawrence Berkeley National Laboratory, 1 Cyclotron Road, Berkeley CA, USA 94720}

\begin{abstract}
\end{abstract}
\begin{keyword}
\end{keyword}
\end{frontmatter}

\section{Introduction}
A class of astronomical objects who share a common luminosity are called a standard
candle.  The relative fluxes (or magnitudes) for a set of these objects distributed in space
provide a measurement of their relative luminosity distances. These data, combined with
measurements of the standard candle redshifts, constitute a Hubble
diagram which is used to map the expansion history of the Universe.

Type~Ia supernovae (SNe~Ia) are an example of a standard candle (more precisely a standardizable
candle), whose redshift-magnitude relation provides an accurate expansion history
used to measure the Hubble Constant
\citep{2001ApJ...553...47F} and detect the accelerated expansion of the Universe
\citep{1998AJ....116.1009R, 1999ApJ...517..565P}.  SN~Ia measurements continue
to improve 
\citep{2014A&A...568A..22B} and play a critical role in probing the physics
responsible for the acceleration \citep{2013PhR...530...87W}.
Looking at the present and toward the future, SN~Ia cosmology is
a component of the Dark Energy Survey\footnote{\url{http://www.darkenergysurvey.org}} (DES),
and a science driver in the design of the
Large Synoptic Survey Telescope\footnote{\url{http://www.lsst.org}} (LSST)
and the Wide-Field InfrarRed Survey Telescope
\citep{2015arXiv150303757S}.

The information necessary to construct a Hubble diagram consists of, per object,
its classification as a
member of the standard candle class, its redshift, and its flux.  For SNe~Ia, the flux at
peak brightness is typically determined from multi-band light curves obtained from repeated
photometric observations. Traditionally, spectroscopic data is used to classify the transient
as SN~Ia through the lack of hydrogen and the presence of SiII in its spectrum,
and determine its redshift either directly through the transient spectrum or that of its host
galaxy. While the classification and redshift could be determined from the multi-band
light curves and colors of the host galaxy, the accuracy is insufficient for precision
cosmology \citep{2011ApJ...738..162S}.  Consequently spectroscopic follow-up
has played a critical role in cosmological supernova surveys.

While the large focal planes of the Dark Energy Camera (3 square degrees)
and the LSST (9.6 square degrees) open the
possibility for the photometric discovery and observation of thousands to millions
of SNe~Ia out to $z\sim1$, a corresponding multiplex advantage has not been
achieved for the 10-m
class telescopes used for the spectroscopic classification of faint discoveries:
it is unfeasible to obtain a spectrum for all SN~Ia candidates with well-measured
photometry.

Photometric SN analysis has been developed
in the hope of taking advantage of the statistical weight imported by the
availability of light curves
for large numbers of supernovae, despite their lack of spectroscopy.
Such analyses have been applied to SDSS-II SN data \citep{2012ApJ...752...79H,
2013ApJ...763...88C}, and are planned for
DES \citep{2012ApJ...753..152B} and LSST \citep{2012arXiv1211.0310L}.

@@

\section{Simple Example}
\subsection{Model}
\label{example:sec}
A simplified version of the above model,
\begin{equation}
p(\mathit{ADU}, {{T}}_S,{{z}}_S, \theta_G|  \Omega_M, w, \theta_r,\alpha_{Ia},\sigma_{Ia}, \alpha_{\mathit{non-Ia}},\sigma_{\mathit{non-Ia}}),
\end{equation}
is examined in this section and is sketched in Figure~\ref{toypgm:fig}.
The measurements are
\begin{itemize}
\item $\mathit{ADU}$: the measured counts at peak brightness;
\item ${{T}}_S$, ${{z}}_S$: the spectroscopic type and redshift for a subset of transients;
\item $\theta_{G1}$, $\theta_{G2}$: the redshifts of the identified host and neighbor
galaxies.
\end{itemize}
The model parameters are
\begin{itemize}
\item $\Omega_M$, $w$: the parameters of a flat dark-energy cosmology;
\item $\theta_{r1}$, $\theta_{r2}$:  for two redshifts the fraction of transients that are SN~Ia;
\item $\alpha_{Ia}$, $\sigma_{Ia}$, $\alpha_{\mathit{non-Ia}}$, $\sigma_{\mathit{non-Ia}}$:
the luminosity (times $4\pi$) and intrinsic dispersion for SNe~Ia and non-Ia at peak.
Light-curve shape and color are not considered in this simple model.
\end{itemize}

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=5in]{/Users/akim/project/abc/results/toy_pgm.pdf} 
   \caption{Probabilistic Graphical Model for the simple SN~Ia analysis. The
   parameters include the cosmological parameters $\Omega_M$ and $w$,
   the luminosities $\alpha$ and intrinsic dispersions $\sigma$ of SNe~Ia and non-Ia's,
   and the relative rate parameters $\theta_{r1}$ and $\theta_{r2}$.  Latent parameters
   include the type $T$ and redshifts of the host and neighboring galaxies $z$ and $z_N$
   for each transient.  The  parameters for distance modulus $\mu$ and
   luminosity $L$ are fixed by the other parameters.  The observables include the measured
   flux $\mathit{ADU}$, spectroscopic type and redshift $T_S$ and $z_S$, and
   the inferred host and neighbor redshifts $\theta_{G1}$ and $\theta_{G2}$.
   \label{toypgm:fig}}
\end{figure}

There are latent model parameters for each transient used internally
by the model:
\begin{itemize}
\item $z$: the cosmological redshifts.
\item $T$: the type, SN~Ia or non-Ia.
\end{itemize}

@@

\subsection{Analysis of Simulated Data}
The simulation of data from the model presented in \S\ref{example:sec}
and their analysis is implemented in Python and are available
on GitHub\footnote{{git@github.com:AlexGKim/abc.git}}.

@@

\begin{table}
\centering
\begin{tabular}{|c|c|cccc|}
\hline
Fraction Spec&\%-ile &$<w>$ & $w$ mode  &$w$ range & $\Delta w$\\ \hline
\multicolumn{6}{|c|}{SN Ia only}\\ \hline
$1.00$& $0.68$ &$-0.94 \pm 0.09$ &$-0.80$ & $[-1.076, -0.801]$ & $0.275$ \\
$1.00$& $0.90$ &$-0.94 \pm 0.09$ &$-0.80$ & $[-1.166, -0.718]$ & $0.448$ \\
$1.00$& $0.95$ &$-0.94 \pm 0.09$ &$-0.80$ & $[-1.208, -0.682]$ & $0.525$ \\
 \hline
\multicolumn{6}{|c|}{Model 1}\\ \hline
$1.00$ &$0.68$ &$-1.06 \pm 0.09$ &$-0.91$ & $[-1.196, -0.914]$ & $0.281$ \\
$1.00$ &$0.90$ &$-1.06 \pm 0.09$ &$-0.91$ & $[-1.292, -0.826]$ & $0.466$ \\
$1.00$ &$0.95$ &$-1.06 \pm 0.09$ &$-0.91$ & $[-1.340, -0.784]$ & $0.556$ \\
$0.60$ &$0.68$ &$-1.08 \pm 0.10$ &$-0.93$ & $[-1.227, -0.933]$ & $0.293$ \\
$0.60$ &$0.90$ &$-1.08 \pm 0.10$ &$-0.93$ & $[-1.331, -0.836]$ & $0.495$ \\
$0.60$ &$0.95$ &$-1.08 \pm 0.10$ &$-0.93$ & $[-1.385, -0.787]$ & $0.598$ \\
$0.20$ &$0.68$ &$-1.10 \pm 0.10$ &$-0.94$ & $[-1.249, -0.943]$ & $0.305$ \\
$0.20$ &$0.90$ &$-1.10 \pm 0.10$ &$-0.94$ & $[-1.354, -0.851]$ & $0.502$ \\
$0.20$ &$0.95$ &$-1.10 \pm 0.10$ &$-0.94$ & $[-1.408, -0.810]$ & $0.597$ \\
$0.00$ &$0.68$ &$-1.07 \pm 0.10$ &$-0.91$ & $[-1.218, -0.909]$ & $0.309$ \\
$0.00$ &$0.90$ &$-1.07 \pm 0.10$ &$-0.91$ & $[-1.320, -0.812]$ & $0.508$ \\
$0.00$ &$0.95$ &$-1.07 \pm 0.10$ &$-0.91$ & $[-1.372, -0.770]$ & $0.602$ \\
 \hline
\multicolumn{6}{|c|}{Model 2}\\ \hline
$1.00$ &$0.68$ &$-0.94 \pm 0.09$ &$-0.80$ & $[-1.078, -0.804]$ & $0.274$ \\
$1.00$ &$0.90$ &$-0.94 \pm 0.09$ &$-0.80$ & $[-1.168, -0.720]$ & $0.447$ \\
$1.00$ &$0.95$ &$-0.94 \pm 0.09$ &$-0.80$ & $[-1.214, -0.680]$ & $0.534$ \\
$0.60$ &$0.68$ &$-1.04 \pm 0.10$ &$-0.88$ & $[-1.182, -0.883]$ & $0.298$ \\
$0.60$ &$0.90$ &$-1.04 \pm 0.10$ &$-0.88$ & $[-1.285, -0.791]$ & $0.494$ \\
$0.60$ &$0.95$ &$-1.04 \pm 0.10$ &$-0.88$ & $[-1.332, -0.749]$ & $0.583$ \\
$0.20$ &$0.68$ &$-1.10 \pm 0.10$ &$-0.94$ & $[-1.247, -0.936]$ & $0.311$ \\
$0.20$ &$0.90$ &$-1.10 \pm 0.10$ &$-0.94$ & $[-1.359, -0.840]$ & $0.518$ \\
$0.20$ &$0.95$ &$-1.10 \pm 0.10$ &$-0.94$ & $[-1.413, -0.798]$ & $0.614$ \\
$0.00$ &$0.68$ &$-1.11 \pm 0.10$ &$-0.95$ & $[-1.266, -0.948]$ & $0.317$ \\
$0.00$ &$0.90$ &$-1.11 \pm 0.10$ &$-0.95$ & $[-1.375, -0.850]$ & $0.525$ \\
$0.00$ &$0.95$ &$-1.11 \pm 0.10$ &$-0.95$ & $[-1.428, -0.808]$ & $0.621$ \\
\hline
\end{tabular}
\caption{Mode and credible intervals of $w$ for only the SNe~Ia, Model 1, and Model 2 for varying
fractions of spectroscopic coverage. \label{seed2:tab}}
\end{table}

@@

When there is 100\% spectroscopic coverage the   $\Omega_M$--$w$  posterior accommodates the input 
parameters, as shown  in
Figure~\ref{contour1:fig}.
A salient feature that contrasts with Model 1 is 
seen in the modes for $w$ listed in Table~\ref{seed2:tab}: biases
larger than the statistical uncertainties appear with decreasing
spectroscopic coverage.  Note that the Hubble diagrams
in these analyses are almost identical: there are redshift misassignments
for 2\% of the transients that lack spectroscopy.

@@

Some initial conclusions.
\begin{itemize}
\item Knowing the underlying model that produces data, you should
get results statistically consistent with the truth.  A tautology?
\item Typing is important when the non-Ia model is not accurate.
\item Better than only using SNe~Ia with spectra.
\end{itemize}

@@

\bibliographystyle{elsarticle-num} 
\bibliography{/Users/akim/Documents/alex}
